---
title: "Estimation"
vignette: >
  %\VignetteIndexEntry{Estimation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
format: 
  html:
    toc: true
toc-depth: 4
number-sections: true
number-depth: 4   
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
#| message: false
#| warning: false
library(dplyr)
library(tidyr)
library(gt)
library(ggplot2)
library(OECDsppps)
library(pricelevels)
library(data.table)
```

# Overview {.unnumbered}

**The estimation** of subnational PPPs (sPPPs) starts with the item-level prices that are progressively aggregated to higher levels. In addition to aggregate subnational price indices, sub-indices, for example, at the COICOP division level, can highlight more granular regional price level differences. This process aligns with current recommendations; see @icp2021, @worldbank2013 and @europeanunionEurostatOECDMethodologicalManual2024 for more information.

## Estimation steps {.unnumbered}

The estimation steps are:

1.  [Estimation of basic headings using item-level prices](#sec-step1), where price data are aggregated up to the level of basic headings, generally without the use of expenditure weights (unless such information is available, for instance, when retail scanner data provide detailed transaction-level records).

2.  [Estimation of higher level aggregates using basic heading indices](#sec-step2) to higher levels of the classification hierarchy, at which point household expenditure data are accessible and can be applied as weighting factors

## Estimation methods {.unnumbered}

The choice of estimation method depends on the availability of data and the analytical objectives of the subnational PPP exercise. When the aim is to ensure cross-country comparability and to exploit micro-level price information, the **Country-Product-Dummy (CPD) - Gini-Éltetö-Köves-Szulc (GEKS)** approach offers a flexible framework for estimating basic heading indices [@icp2021].

In contrast, the **Eurostat OECD method (Jevons-GEKS)** methodology employed at the national level imposes more stringent data requirements. Such requirements may be more difficult to meet in the context of deriving subnational PPPs based on existing microdata, particularly regarding the representativeness of individual products across all regions [@europeanunionEurostatOECDMethodologicalManual2024].

The estimation procedure in this vignette follows the CPD-GEKS approach and highlights its similarities and differences with the Jevons-GEKS approach whenever instructive. For a more comprehensive discussion on price indices see @worldbank2013 and @europeanunionEurostatOECDMethodologicalManual2024.

## Data used {.unnumbered}

Data used in this vignette is taken from official UK microdata from the United Kingdom Office for National Statistics (ONS). Similar data was recently used in @hearne2025 and is publicly available:

-   `uk_cpi()` is a snipped of the [UK CPI microdata](https://www.ons.gov.uk/economy/inflationandpriceindices/datasets/consumerpriceindicescpiandretailpricesindexrpiitemindicesandpricequotes) containing two products: White sliced loaf branded 750 grams (COICOP 1010103) and carpenter hourly rate (COICOP 410518).

-   `uk_hhx()` is a snipped of the regional UK household expenditure data ...🚧 work in progress....

------------------------------------------------------------------------

# Estimation of basic headings using item-level prices {#sec-step1}

## Overview

The CPD method is a regression-based approach for estimating price parities. The underlying statistical model is

$$
p_{ij} = PPP_j \times p_i \times \epsilon_{ij}
$$ {#eq-ppp1}

where $PPP_j$ is the purchasing power parity of an arbitrary region $j$, ($r = 1,...,j,...,R$), $p_i$ is the average regional price of an arbitrary commodity $i$, ($n = 1, ..., i, ... N$), and $\epsilon_{ij}$ is a independently and identically distributed random variable.[^1] Taking logs of @eq-ppp1 yields

[^1]: Model @eq-ppp1 is not identified and requires parametrisation before it can be estimated.

$$
\begin{aligned}
ln p_{ij} & = ln PPP_j + ln p_i  + ln \epsilon_{ij} \\
& = \alpha_j + \gamma_i + ln \varepsilon_{ij}
\end{aligned}
$$ {#eq-ppp2}

where $\alpha_j$ is the the price level of region $j$ relative to all other regions in the comparison. $\alpha_j$ can also be expressed relative to a reference region, for example, the national price level. Then, $\alpha_j$  represents the subnational purchasing power parity of region $j$ given by $\hat{PPP}_j = exp(\hat{\alpha}_j)$.

## Estimation

The CPD model in @eq-ppp2 may be interpreted as a fixed-effects specification, in which country effects yield estimates of subnational purchasing power parities, while commodity-specific effects generate estimates of subnational price levels. The model can be written as a regression equation in which all explanatory variables take the form of dummy indicators for each region and commodity.

$$
\begin{aligned}
ln p_{ij}  = & \alpha_1 D_1 + ... + \alpha_j D_j + ... +\alpha_R D_R + \\
& \eta_1 \mathcal{D}_1 + ... + \eta_i \mathcal{D}_i + ... + \eta_N \mathcal{D}_N + \varepsilon_{ij}
\end{aligned}
$$ {#eq-ppp3}

and $\varepsilon_{ij}$ are independently and identically (normally) distributed with a zero mean and variance $\sigma^2$, that is, $\varepsilon_{ij} \sim N(0, \sigma^2)$. The variables of interest, $PPP_j$, can be estimated through the parameters $\hat{\alpha}_j$  using ordinary least squares (OLS); see [Section 1.1.3 Implementation](#sec-implementation).



## Implementation {#sec-implementation}


This section describes the implementation of the CPD using different approaches
from simple cross-tabulations, standard OLS as well as using the 
`pricelevels` package [@pricelevels] and finally the function `estim_cpd()` of
this package using examples [1](#sec-example1) and [2](#sec-example2).

Examples [3](#sec-example3) to [6](#sec-example6) discuss further functionalities of
`estim_cpd()`.

------------------------------------------------------------------------

### Example 1: One product, two regions {#sec-example1 .unnumbered}

### Using cross-tabulations

```{r}
# Data
df1 <- data.table(
  region = as.factor(c(1, 2, 1, 2)),
  product = as.factor(c(1, 1, 1, 1)),
  price = c(25, 28, 23, 26)
)
```

In the CPD regression model, the intercept corresponds to the cross-regional average.

```{r}
# Calculate cross-regional price average
df1 |>
  as_tibble() |>
  summarise(mean(price))
log(25.5)
```

And the coefficient estimate is the price ratio of the average regional prices.

```{r}
# Calculate regional price averages and price relative
df1 |>
  as_tibble() |>
  group_by(region) |>
  summarise(mean(price))
27 / 24
```

### Using `pricelevels`

The same results can be obtained using `cpd()` from the `pricelevels` package.

```{r}
#| message: false
#| warning: false
# With pricelevels - estimation with respect to regional average
df1[, cpd(p = price, r = region, n = product, q = NULL, base = NULL)]
1.060660 / 0.942809

# With pricelevels - estimation with respect to region 1
df1[, cpd(p = price, r = region, n = product, q = NULL, base = "1")]

# With pricelevels - estimation output
df1[, cpd(
  p = price, r = region, n = product, q = NULL, base = NULL,
  simplify = FALSE
)]
exp(-0.05889)
```

### Using standard OLS

The same can be achieved with a simple OLS regression

```{r}
#| message: false
#| warning: false
# With OLS
pdata <- df1

# Model: add intercept to for price levels relative to base
cpd_mod <- log(price) ~ region + 1

# Transformation: equiv to "mean centring" of continuous predictor: shifting
## dummy encoding to -1/1 intercept is mean across all prices.
## Region1 is the 'main effect', i.e., the difference between levels of a given
## factor (region) across all other factors.
contrasts(x = pdata$region) <- contr.sum(levels(pdata$region))
colnames(contrasts(x = pdata$region)) <- levels(pdata$region)[-nlevels(pdata$region)]

# OLS regression
out <- lm(formula = cpd_mod, data = pdata)
out
exp(dummy.coef(out)[["region"]])
exp(3.23617 - 0.05898)
exp(3.23617 + 0.05898)
```

### Example 2: Two products, two regions {#sec-example2 .unnumbered}

The procedure is identical for the more general case.

```{r}
#| message: false
#| warning: false
# Data
df2 <- data.table(
  region = as.factor(c(1, 2, 1, 2)),
  product = as.factor(c(1, 1, 2, 2)),
  price = c(25, 28, 23, 26)
)
```

### Using `pricelevels`

```{r}
#| message: false
#| warning: false
# With pricelevels} ------
## Estimation with respect to regional average
df2[, cpd(p = price, r = region, n = product, q = NULL, base = NULL)]
1.060757 / 0.942723

## Estimation with respect to region 1
df2[, cpd(p = price, r = region, n = product, q = NULL, base = "1")]

## Estimation output
df2[, cpd(
  p = price, r = region, n = product, q = NULL, base = NULL,
  simplify = FALSE
)]
exp(-0.05898)
```

### Using standard OLS

```{r}
#| message: false
#| warning: false
# OLS ------
## Data
pdata <- df2
## Model
cpd_mod <- log(price) ~ product + region - 1

## Transformation: equiv to "mean centring"
contrasts(x = pdata$region) <- contr.sum(levels(pdata$region))
colnames(contrasts(x = pdata$region)) <- levels(pdata$region)[-nlevels(pdata$region)]

## OLS regression
out <- lm(formula = cpd_mod, data = pdata)
out
exp(dummy.coef(out)[["region"]])
```

------------------------------------------------------------------------

### Integration in \pkg{OECDsppps}:  `estim_cpd()`

\pkg{OECDsppps} provides the function `estim_cpd()` for CPD estimation. The 
function provides nummerically identical results as the previously discussed
estimations and provides further functionalities; see the examples below as
well as the documentation of `estim_cpd()` for more information.

#### Example 3: Generic - Multiple products, and regions, with and without weights {#sec-example3 .unnumbered}


```{r}
#| message: false
#| warning: false
# Generate data with pricelevels -------
set.seed(123)
R <- 5 # number of regions
B <- 5 # number of product groups
N <- 5 # number of products
dt1 <- pricelevels::rdata(R = R, B = B, N = N)
```

CPD with no weights using `cpd()` in `pricelevels` [@pricelevels] and `estim_cpd()`.

```{r}
#| message: false
#| warning: false
# Estimating sPPPs with `pricelevels`, no weights --------
dt1[, cpd(p = price, r = region, n = product)]

# Estimating sPPPs with `estim_cpd()`, no weights ---------
dt1 |>
  estim_cpd(
    region = "region",
    product = "product",
    price = "price"
  ) |>
  pull("sPPP")
```

CPD with weights using `cpd()` in `pricelevels` [@pricelevels] and `estim_cpd()`.

```{r}
# Estimating sPPPs with `pricelevels`, with weights --------
dt1[, cpd(p = price, r = region, n = product, w = weight)]

# Estimating sPPPs with `estim_cpd()`, with weights ---------
dt1 |>
  estim_cpd(
    region = "region",
    product = "product",
    price = "price",
    weights_cpd = "weight"
  ) |>
  pull("sPPP")
```


#### Example 4: Complete regression output {#sec-example4 .unnumbered}


The function `estim_cpd()` also has the option to export extended regression output of the CPD model with argument `output = "Full"`, which summarises the key information of the estimate CPD model: It provides the 'Regression output`
as well as the individual 'Residuals' of the CPD regression. 

Information in the extended regression output is used to support the validation of CPD-based subnational PPPs at the basic-heading level; see [Validation](https://amannj.github.io/OECDsPPPs/articles/Validation.html#sec-tobh) vignette.

```{r}
# Estimating sPPPs with `estim_cpd()` ---------
full_est <- dt1 |>
  estim_cpd(
    region = "region",
    product = "product",
    price = "price",
    output = "Full"
  )

## Regression output
full_est[["Regression output"]] |>
  gt() |>
  fmt_number(decimals = 2) |>
  sub_missing(missing_text = "")

## Residuals
full_est[["Residuals"]] |>
  head() |>
  gt() |>
  fmt_number(decimals = 4) |>
  sub_missing(missing_text = "")
```



#### Example 5: Duplicate region-product price pairs defaults {#sec-example5 .unnumbered}

By default, `estim_cpd()` aggregates the price quotes up to region-product pairs 
using unweighted means whenever there are duplicate region-product pairs found in data 
and no weights provided. This is identical to the bahaviour of `cpd()` in `pricelevels`

```{r}
# Take UK CPI microdata containing duplicate region-product pairs ---------
red <- uk_cpi |>
  filter(Year == "2018") |>
  select(
    region = "Region",
    product = "Product code",
    price = "Reference quantity price"
  ) |>
  mutate(
    region = as.factor(region),
    product = as.factor(product)
  )

# Estimating sPPPs with `estim_cpd()` ---------
red |>
  estim_cpd() |>
  pull("sPPP")

# Estimating sPPPs with `pricelevels` --------
as.data.table(red)[, cpd(p = price, r = region, n = product)]
```


#### Example 6: Duplicate region-product price pairs with aggregation weights {#sec-example6 .unnumbered}


`estim_cpd()` also provides the option to add aggregation weight in case duplicate 
region-product pairs found in data through the `weights` argument; see `estim_cpd()` for more information.

```{r}
# Estimating sPPPs with `estim_cpd()`, with aggregation weights ---------
## No weights
red |>
  mutate(w = 1) |>
  estim_cpd(weights = "w") |>
  pull("sPPP")

## Random weights
set.seed(123)
red |>
  mutate(w = runif(nrow(red))) |>
  estim_cpd(weights = "w") |>
  pull("sPPP")
```

#### Example 7: Duplicate region-product price pairs without aggregation {#sec-example7 .unnumbered}


`estim_cpd()` also provides the option to run the CPD method on the raw data, 
that is, keeping duplicate region-product pairs found in the raw data 
by setting `weights = 'raw'` argument; see `estim_cpd()` for more information.

```{r}
red |>
  estim_cpd(weights = "raw") |>
  pull("sPPP")
```

------------------------------------------------------------------------

# Estimation of higher level aggregates using basic heading indices {#sec-step2}

## Overview

**Laspeyres index** is obtained as

$$
sPPP_L^{j,k} = \sum_{n=1}^N  w_n^j \times sPPP_n^{j,k}
$$ {#eq-lasp}

**Paasche index** is obtained as

$$
sPPP_P^{j,k} = \frac{1}{\sum_{n=1}^{N} \frac{w_n^k}{sPPP_n^{j,k}}}
$$ {#eq-paas}

**Fisher index** is obtained as

$$
sPPP_F^{j,k} = \left( sPPP_L^{j,k} \times sPPP_P^{j,k} \right)^{1/2}
$$ {#eq-fish}

**Gini-Éltetö-Köves-Szulc index (GEKS)** is obtained as

$$
sPPP_{GEKS}^{j,k} = \prod_{r=1}^R \left( sPPP_F^{j,r} \times sPPP_F^{r,k} \right)^{1/R}
$$ {#eq-fish}

## Implementation

This section illustrates the calculation of Laspeyres, Paasche, Fisher, and GEKS indices using the `index_laspeyres()`, `index_paasche()`, `index_fisher()`, `index_geks()` functions.

------------------------------------------------------------------------

Generate the basic heading level CPD estimates

```{r}
set.seed(123)

R <- 5 # number of regions
B <- 5 # number of product groups
N <- 5 # number of products

dt1 <- pricelevels::rdata(R = R, B = B, N = N,
                          weights = ~r+n,
                          settings = list(par.sd = c(lnP = 0.1, pi = exp(1), delta = 0.5, error = 0.8))
                          ) %>%
  as_tibble()

# Get the CPD estimate
cpd_oecd <- dt1 %>%
  dplyr::select(group, region, product, price) %>%
  nest(.by = c(group)) %>%
  mutate(cpd_estimation = map(.x = data, ~ .x %>%
                                estim_cpd(region = "region",
                                          product = "product",
                                          price = "price"))) %>%
  unnest(cpd_estimation) %>%
  left_join(dt1[, c("region", "group", "weight")] %>%
              distinct(region, group, .keep_all = TRUE),
            by = c("group", "region"))
```

Calculate the indices using `OECDsppps` functions
```{r}
indices_oecdsppps_df <- cpd_oecd %>%
  reframe(index_laspeyeres(data = .,
                           region = "region",
                           product = "group",
                           ppp_bh = "sPPP",
                           exp_wght = "weight"),
          index_paasche(data = .,
                        region = "region",
                        product = "group",
                        ppp_bh = "sPPP",
                        exp_wght = "weight"),
          index_fisher(data = .,
                       region = "region",
                       product = "group",
                       ppp_bh = "sPPP",
                       exp_wght = "weight"),
          index_geks(data = .,
                     region = "region",
                     product = "group",
                     ppp_bh = "sPPP",
                     exp_wght = "weight")) %>%
  pivot_longer(!c(base_region, region), names_to = "index_type", values_to = "index_value") %>%
  mutate(package = "OECDsppps")
```

Calculate the indices using `pricelevels`
```{r}
cpd_oecd_pl <- cpd_oecd %>%
  dplyr::select(group, region, sPPP, weight) %>%
  as.data.table()

regions_v <- c("1", "2", "3", "4", "5")

laspeyres_pl <- vector("list", 5)
paasche_pl <- vector("list", 5)
fisher_pl <- vector("list", 5)
geks_pl <- vector("list", 5)

names(laspeyres_pl) <- regions_v
names(paasche_pl) <- regions_v
names(fisher_pl) <- regions_v
names(geks_pl) <- regions_v


for (i in seq_along(regions_v)) {
  laspeyres_pl[[i]] <- cpd_oecd_pl[, laspeyres(p = sPPP, r = region, n = group, w = weight, base = c(regions_v[i]))]
  paasche_pl[[i]] <- cpd_oecd_pl[, paasche(p = sPPP, r = region, n = group, w = weight, base = c(regions_v[i]))]
  fisher_pl[[i]] <- cpd_oecd_pl[, fisher(p = sPPP, r = region, n = group, w = weight, base = c(regions_v[i]))]
  geks_pl[[i]] <- cpd_oecd_pl[, geks(p = sPPP, r = region, n = group, w = weight, settings = list(type = "fisher"), base = c(regions_v[i]))]
}


indices_pricelevels_df <- do.call(cbind, laspeyres_pl) %>%
  as_tibble() %>%
  rownames_to_column(var = "region") %>%
  pivot_longer(!region, names_to = "base_region", values_to = "laspeyres_index") %>%
  left_join(do.call(cbind, paasche_pl) %>%
              as_tibble() %>%
              rownames_to_column(var = "region") %>%
              pivot_longer(!region, names_to = "base_region", values_to = "paasche_index"),
            by = c("base_region", "region")) %>%
  left_join(do.call(cbind, fisher_pl) %>%
              as_tibble() %>%
              rownames_to_column(var = "region") %>%
              pivot_longer(!region, names_to = "base_region", values_to = "fisher_index"),
            by = c("base_region", "region")) %>%
  left_join(do.call(cbind, geks_pl) %>%
              as_tibble() %>%
              rownames_to_column(var = "region") %>%
              pivot_longer(!region, names_to = "base_region", values_to = "geks_index"),
            by = c("base_region", "region")) %>%
  pivot_longer(!c(base_region, region), names_to = "index_type", values_to = "index_value") %>%
  mutate(package = "pricelevels")
```

Comparison of `OECDsppps` and `pricelevels` output
```{r}
indices_oecdsppps_df %>%
  bind_rows(indices_pricelevels_df) %>%
  mutate(index_type = str_remove_all(index_type, "_index"),
         base_region = paste("Base region:", base_region)) %>%
  ggplot(aes(x = region, y = index_value, fill = package)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "OECDsppps/pricelevels comparison",
    x = "region",
    y = ""
  ) +
  theme_minimal() +
  theme(legend.position = "top") +
  scale_fill_manual(values = c("#a3bbdd", "#2a4691")) +
  geom_hline(yintercept = 1) +
  facet_wrap(~ base_region + index_type, ncol = 4)
```

------------------------------------------------------------------------

> 🚧 Additional sections remain work in progress.

------------------------------------------------------------------------


# References {.unnumbered}
