---
title: "Estimation"
vignette: >
  %\VignetteIndexEntry{Estimation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
format: 
  html:
    toc: true
toc-depth: 4
number-sections: true
number-depth: 4   
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: setup
#| message: false
#| warning: false
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(tibble)
library(gt)
library(ggplot2)
library(OECDsppps)
library(pricelevels)
library(data.table)
```

# Overview {.unnumbered}

**The estimation** of subnational PPPs (sPPPs) starts with the
item-level prices that are progressively aggregated to higher levels. In
addition to aggregate subnational price indices, sub-indices, for
example, at the COICOP division level, can highlight more granular
regional price level differences. This process aligns with current
recommendations; see @icp2021, @worldbank2013 and
@europeanunionEurostatOECDMethodologicalManual2024 for more information.

## Estimation steps {.unnumbered}

The estimation steps are:

1.  [Estimation of basic headings using item-level prices](#sec-step1),
    where price data are aggregated up to the level of basic headings,
    generally without the use of expenditure weights (unless such
    information is available, for instance, when retail scanner data
    provide detailed transaction-level records).

2.  [Estimation of higher-level aggregates using basic heading
    indices](#sec-step2) to higher levels of the classification
    hierarchy, at which point household expenditure data are accessible
    and can be applied as weighting factors.

To seamlessly combine the two estimation steps, `OECDsppps`, provides
the 'linking function' `estim_index_link()` to impute missing
basic-heading subnational PPPs. A complete sylised workflow is described
in section [Putting it all together](#sec-combined).

## Estimation methods {.unnumbered}

The choice of estimation method depends on the availability of data and
the analytical objectives of the subnational PPP exercise. When the aim
is to ensure cross-country comparability and to exploit micro-level
price information, the **Country-Product-Dummy - Gini-Éltetö-Köves-Szulc
(CPD-GEKS)** approach offers a flexible framework for estimating
basic-heading indices [@icp2021], where the first estiation step to the
basic-heading level is performed using the CPD method, and further
aggregation beyond the basic-heading level is carried out using GEKS.
Basic-heading aggregation with CPD is recommended by the ICP [@icp2021]
as it is better-suited to handling missing price observations in the
underlying price microdata than GEKS [@auer_von_gap_2026].[^1]

[^1]: This is particularly important when deriving PPPs from existing
    price microdata, where not all products are observed in every
    region, rather than from dedicated PPP surveys based on a
    pre-specified list of comparable items. In particular, GEKS
    price-level estimates are intrinsically more sensitive to missing
    observations than those obtained through CPD and therefore
    diagnostic tools for assessing data completeness are especially
    valuable for GEKS-based price-level estimates [@auer_von_gap_2026].

In contrast, the **Eurostat-OECD (Jevons-GEKS)** method employed at the
national level imposes more stringent data requirements. Such
requirements may be more difficult to meet in the context of deriving
subnational PPPs based on existing microdata, particularly regarding the
representativeness of individual products across all regions
[@europeanunionEurostatOECDMethodologicalManual2024].

The estimation procedure in this vignette follows the CPD-GEKS approach
and highlights its similarities and differences with the Jevons-GEKS
approach whenever instructive. For a more comprehensive discussion on
price indices see @worldbank2013 and
@europeanunionEurostatOECDMethodologicalManual2024.

------------------------------------------------------------------------

# Estimation of basic headings using item-level prices {#sec-step1}

## Overview

The CPD method is a regression-based approach for estimating price
parities. The underlying statistical model is

$$p_{ij} = PPP_j \times p_i \times \epsilon_{ij}$$ {#eq-ppp1}

where $PPP_j$ is the purchasing power parity of an arbitrary region $j$,
($r = 1,...,j,...,R$), $p_i$ is the average cross-regional price of the
same arbitrary commodity $i$, ($n = 1, ..., i, ... N$), and
$\epsilon_{ij}$ is an independently and identically distributed random
variable.[^2] Taking logs of @eq-ppp1 yields

[^2]: Model @eq-ppp1 is not identified and requires parametrisation
    before it can be estimated.

$$\begin{aligned}
ln p_{ij} & = ln PPP_j + ln p_i  + ln \epsilon_{ij} \\
& = \alpha_j + \gamma_i + ln \varepsilon_{ij}
\end{aligned}$$ {#eq-ppp2}

where $\alpha_j$ is the price level of region $j$ relative to all other
regions in the comparison. $\alpha_j$ can also be expressed relative to
a reference region, for example, the national price level. Then,
$\alpha_j$ represents the subnational purchasing power parity of region
$j$ given by $\hat{PPP}_j = exp(\hat{\alpha}_j)$.

## Estimation

The CPD model in @eq-ppp2 may be interpreted as a fixed-effects
specification, in which the region effects yield estimates of
subnational Purchasing Power Parities, while commodity-specific effects
generate estimates of subnational price levels. The model can be written
as a regression equation in which all explanatory variables take the
form of dummy indicators for each region and commodity.

$$\begin{aligned}
ln p_{ij}  = & \alpha_1 D_1 + ... + \alpha_j D_j + ... +\alpha_R D_R + \\
& \eta_1 \mathcal{D}_1 + ... + \eta_i \mathcal{D}_i + ... + \eta_N \mathcal{D}_N + \varepsilon_{ij}
\end{aligned}$$ {#eq-ppp3}

and $\varepsilon_{ij}$ are independently and identically (normally)
distributed with a zero mean and variance $\sigma^2$, that is,
$\varepsilon_{ij} \sim N(0, \sigma^2)$. The variables of interest,
$PPP_j$, can be estimated through the parameters $\hat{\alpha}_j$ using
ordinary least squares (OLS).

## Implementation {#sec-implementation}

This section describes the implementation of the CPD using different
approaches from simple cross-tabulations, standard OLS as well as using
the `pricelevels` package [@pricelevels] and finally the function
`estim_cpd()` of this package using examples [1](#sec-example1) and
[2](#sec-example2).

Examples [3](#sec-example3) to [6](#sec-example6) discuss further
functionalities of `estim_cpd()`.

------------------------------------------------------------------------

### Example 1: One product, two regions {#sec-example1 .unnumbered}

### Using cross-tabulations

```{r}
# Data
df1 <- data.table(
  region = as.factor(c(1, 2, 1, 2)),
  product = as.factor(c(1, 1, 1, 1)),
  price = c(25, 28, 23, 26)
)
```

In the CPD regression model, the intercept corresponds to the
cross-regional average.

```{r}
# Calculate cross-regional price average
df1 |>
  as_tibble() |>
  summarise(mean(price))
log(25.5)
```

The coefficient estimate is the price ratio of the average regional
prices.

```{r}
# Calculate regional price averages and price relative
df1 |>
  as_tibble() |>
  group_by(region) |>
  summarise(mean(price))
27 / 24
```

### Using `pricelevels`

The same results can be obtained using `cpd()` from the `pricelevels`
package.

```{r}
#| message: false
#| warning: false
# With pricelevels - estimation with respect to regional average
df1[, cpd(p = price, r = region, n = product, q = NULL, base = NULL)]
1.060660 / 0.942809

# With pricelevels - estimation with respect to region 1
df1[, cpd(p = price, r = region, n = product, q = NULL, base = "1")]

# With pricelevels - estimation output
df1[, cpd(
  p = price, r = region, n = product, q = NULL, base = NULL,
  simplify = FALSE
)]
exp(-0.05889)
```

### Using standard OLS

The same can be achieved with a simple OLS regression

```{r}
#| message: false
#| warning: false
# With OLS
pdata <- df1

# Model: add intercept to for price levels relative to base
cpd_mod <- log(price) ~ region + 1

# Transformation: equiv to "mean centring" of continuous predictor: shifting
## dummy encoding to -1/1 intercept is mean across all prices.
## Region1 is the 'main effect', i.e., the difference between levels of a given
## factor (region) across all other factors.
contrasts(x = pdata$region) <- contr.sum(levels(pdata$region))
colnames(contrasts(x = pdata$region)) <- levels(pdata$region)[-nlevels(pdata$region)]

# OLS regression
out <- lm(formula = cpd_mod, data = pdata)
out
exp(dummy.coef(out)[["region"]])
exp(3.23617 - 0.05898)
exp(3.23617 + 0.05898)
```

### Example 2: Two products, two regions {#sec-example2 .unnumbered}

The procedure is identical for the more general case.

```{r}
#| message: false
#| warning: false
# Data
df2 <- data.table(
  region = as.factor(c(1, 2, 1, 2)),
  product = as.factor(c(1, 1, 2, 2)),
  price = c(25, 28, 23, 26)
)
```

### Using `pricelevels`

```{r}
#| message: false
#| warning: false
# With pricelevels ------
## Estimation with respect to regional average
df2[, cpd(p = price, r = region, n = product, q = NULL, base = NULL)]
1.060757 / 0.942723

## Estimation with respect to region 1
df2[, cpd(p = price, r = region, n = product, q = NULL, base = "1")]

## Estimation output
df2[, cpd(
  p = price, r = region, n = product, q = NULL, base = NULL,
  simplify = FALSE
)]
exp(-0.05898)
```

### Using standard OLS

```{r}
#| message: false
#| warning: false
# OLS ------
## Data
pdata <- df2
## Model
cpd_mod <- log(price) ~ product + region - 1

## Transformation: equiv to "mean centring"
contrasts(x = pdata$region) <- contr.sum(levels(pdata$region))
colnames(contrasts(x = pdata$region)) <- levels(pdata$region)[-nlevels(pdata$region)]

## OLS regression
out <- lm(formula = cpd_mod, data = pdata)
out
exp(dummy.coef(out)[["region"]])
```

------------------------------------------------------------------------

### Integration in \pkg{OECDsppps} using `estim_cpd()`

\pkg{OECDsppps} provides the function `estim_cpd()` for CPD estimation.
The function provides nummerically identical results as the previously
discussed estimations and provides further functionalities; see the
examples below as well as the documentation of `estim_cpd()` for more
information.

#### Example 3: Multiple products and regions, with and without weights {#sec-example3 .unnumbered}

CPD with no weights using `cpd()` in `pricelevels` [@pricelevels] and
`estim_cpd()`.

```{r}
#| message: false
#| warning: false
# Estimating sPPPs with `pricelevels`, no weights --------
as.data.table(sampledata_prices)[, cpd(p = price, r = region, n = product)]

# Estimating sPPPs with `estim_cpd()`, no weights ---------
sampledata_prices |>
  estim_cpd(
    region = "region",
    product = "product",
    price = "price"
  ) |>
  pull("sPPP")
```

CPD with weights using `cpd()` in `pricelevels` [@pricelevels] and
`estim_cpd()`.

```{r}
# Combining sample price quotes and weights into one data object
sampledata <- sampledata_prices |>
  left_join(sampledata_weights)

# Estimating sPPPs with `pricelevels`, with weights --------
as.data.table(sampledata)[, cpd(p = price, r = region, n = product, w = exp_wght)]

# Estimating sPPPs with `estim_cpd()`, with weights ---------
sampledata |>
  estim_cpd(
    region = "region",
    product = "product",
    price = "price",
    weights_cpd = "exp_wght"
  ) |>
  pull("sPPP")
```

#### Example 4: Complete regression output {#sec-example4 .unnumbered}

The function `estim_cpd()` also has the option to export extended
regression output of the CPD model with argument `output = "Full"`,
which summarises the key information of the estimated CPD model: It
provides the 'Regression output' as well as the individual 'Residuals'
of the CPD regression.

Information in the extended regression output is used to support the
validation of CPD-based subnational PPPs at the basic-heading level; see
[Validation](https://amannj.github.io/OECDsppps/articles/Validation.html#sec-tobh)
vignette.

```{r}
# Estimating sPPPs with `estim_cpd()` ---------
full_est <- sampledata_prices |>
  estim_cpd(
    region = "region",
    product = "product",
    price = "price",
    output = "Full"
  )

## Regression output
full_est[["Regression output"]] |>
  gt() |>
  fmt_number(decimals = 2) |>
  sub_missing(missing_text = "")

## Residuals
full_est[["Residuals"]] |>
  head() |>
  gt() |>
  fmt_number(decimals = 4) |>
  sub_missing(missing_text = "")
```

#### Example 5: Duplicate region-product price pairs defaults {#sec-example5 .unnumbered}

By default, `estim_cpd()` aggregates the price quotes up to
region-product pairs using unweighted means whenever there are duplicate
region-product pairs found in data and no weights provided. This is
identical to the behaviour of `cpd()` in `pricelevels` .

```{r}
# Estimating sPPPs with `estim_cpd()` ---------
sampledata_multi_period |>
  estim_cpd(
    region = "region",
    product = "product",
    price = "price"
  ) |>
  pull(sPPP)

# Estimating sPPPs with `pricelevels` --------
as.data.table(sampledata_multi_period)[, cpd(p = price, r = region, n = product)]
```

#### Example 6: Duplicate region-product price pairs with aggregation weights {#sec-example6 .unnumbered}

`estim_cpd()` also provides the option to add aggregation weight in case
duplicate region-product pairs found in data through the `weights`
argument; see `estim_cpd()` for more information.

```{r}
# Estimating sPPPs with `estim_cpd()`, with aggregation weights ---------
## No weights
sampledata_prices |>
  mutate(w = 1) |>
  estim_cpd(weights = "w") |>
  pull("sPPP")

## Random weights
set.seed(123)
sampledata_prices |>
  mutate(w = runif(nrow(sampledata_prices))) |>
  estim_cpd(weights = "w") |>
  pull("sPPP")

## Generic sample weights
sampledata_prices |>
  left_join(sampledata_weights) |>
  estim_cpd(weights = "exp_wght") |>
  pull("sPPP")
```

#### Example 7: Duplicate region-product price pairs without aggregation {#sec-example7 .unnumbered}

`estim_cpd()` also provides the option to run the CPD method on the raw
data, that is, keeping duplicate region-product pairs found in the raw
data by setting the `weights` argument to `'raw'`; see `estim_cpd()` for
more information.

```{r}
sampledata_prices |>
  mutate(w = runif(nrow(sampledata_prices))) |>
  estim_cpd(weights = "raw") |>
  pull("sPPP")
```

------------------------------------------------------------------------

# Estimation of higher-level aggregates using basic heading indices {#sec-step2}

## Overview

Aggregation beyond the basic-heading level uses the
Gini-Éltetö-Köves-Szulc index (GEKS) and is implemented using function
`index_geks()`.

The GEKS method is constructed based on Fisher, Laspeyres and Paasche
price indices. These index calculations are integrated in
`index_geks()`, but the separate indices can also be calculated
separately using the functions `index_laspeyres()`, `index_paasche()`
and `index_fisher()`.

## Indices {#sec-step2Indices}

### Gini-Éltetö-Köves-Szulc index (GEKS) {#sec-geks .unnumbered}

The Gini-Éltetö-Köves-Szulc index (GEKS) method is recommended for
aggregating above the basic-heading levels for international and
interregional comparisons, as it satisfies the necessary properties for
multilateral comparisons; see @icp2021, @worldbank2013 and
@europeanunionEurostatOECDMethodologicalManual2024 for more information.

Subnational PPPs for region $k$ with reference to region $j$ are
calculated using the GEKS as

$$sPPP_{GEKS}^{j,k} = \prod_{r=1}^R \left( sPPP_F^{j,r} \times sPPP_F^{r,k} \right)^{1/R}$$

and correspond to the geometric average of the [Fisher
indices](#sec-fisher) of all direct comparisons between region $j$ and
region $k$, and indirect comparisons across all regions $R$
($r=1,\dots,j,k,\dots,R$).

The Gini-Éltetö-Köves-Szulc index (GEKS) is implemented through function
`index_geks()`.

### Fisher index {#sec-fisher .unnumbered}

For the two arbitrary regions $j$ and $k$, the Fisher index calculates
the subnational PPPs for the geometric mean of the weighted basic
heading $b$, $b = 1,\dots, B$, of the [Laspeyres index](#sec-geks),
which holds quantities in region $j$ constant and allows prices to
change, and the [Paasche index](#sec-geks), which holds quantities in
region $k$ constant and allows prices to change.

$$sPPP_F^{j,k} = \left( sPPP_L^{j,k} \times sPPP_P^{j,k} \right)^{1/2}$$
{#eq-fish}

The Fisher index is calculated based on the results estimated at the
basic heading level, where $p_{jk}$ and $e_{jk}$, respectively, denote
the price and associated expenditure for an arbitrary basic heading $b$
and region $r$, and the implied quantity, $q_{br}$, is given by
$q_{br} =  \frac{e_{br} }{ p_{br}}$:

$$sPPP_F^{j,k} = \left[ \underbrace{ \frac{\sum_{b = 1}^B p_{bk}q_{bj}}{\sum_{b = 1}^B p_{bj}q_{bj}} }_{sPPP_L^{j,k}} \times \underbrace{ \frac{\sum_{b = 1}^B p_{bk}q_{bk}}{\sum_{b = 1}^B p_{bj}q_{bk}} }_{sPPP_P^{j,k}}   \right]^{1/2}$$

The Fisher index is implemented through function `index_fisher()`.

### Laspeyres index {#sec-laspeyres .unnumbered}

The Laspeyres index evaluates how much the given consumer basket in base
region $j$ would cost if the basket were priced at partner region $k$'s
prices.

The Laspeyres is calculated as

$$sPPP_L^{j,k} = \sum_{n=1}^N  w_n^j \times sPPP_n^{j,k}$$ {#eq-lasp}

The Laspeyres index is implemented using function `index_laspeyres()`.

### Paasche index {#sec-paasche .unnumbered}

The same logic is mirrored with the Paasche index, where region $k$'s
basket is evaluated and compared with region $j$'s price. The Paasche
index is calculated as

$$sPPP_P^{j,k} = \frac{1}{\sum_{n=1}^{N} \frac{w_n^k}{sPPP_n^{j,k}}}$$
{#eq-paas}

The Paasche index is implemented using function `index_paasche()`.

## Implementation

This section illustrates the calculation of Laspeyres, Paasche, Fisher,
and GEKS indices using the `index_laspeyres()`, `index_paasche()`,
`index_fisher()`, `index_geks()` in `OECDsppps`.

It also illustrates the calculations of the same indices in the
`pricelevels` package [@pricelevels].

For this, some generic CPD estimates are generated first.

```{r}
# Calculate generic CPD estimates
cpd_oecd <- sampledata_prices %>%
  dplyr::select(heading, region, product, price) %>%
  group_by(heading) %>%
  group_modify(~ {
    estim_cpd(.x,
      region = "region",
      product = "product",
      price = "price",
      output = "sPPP"
    )
  }) %>%
  ungroup() %>%
  left_join(
    sampledata_weights[, c("region", "heading", "exp_wght")] %>%
      distinct(region, heading, .keep_all = TRUE),
    by = c("heading", "region")
  )
```

### Price indices in `OECDsppps`

The different price indices can be calculated through the functions
`index_laspeyres()`, `index_paasche()`, `index_fisher()` and
`index_geks()`, and take identical input arguments.

Note that the `reframe()` function, from package `dplyr` [@refdplyr], in
the code chunk below creates a new data frame around the returned set of
index estimates. This is useful if the separate indices should be
returned together; however, it is not necessary to calculate all these
indices separately to implement the GEKS method, as `index_geks()`
carries out all necessary calculation steps.

In other words, `index_geks()` nests the functions `index_fisher()`,
`index_laspeyres()` and `index_paasche()`, respectively, as discussed in
the [Indices](#sec-step2Indices) section.

```{r}
indices_oecdsppps_df <- cpd_oecd %>%
  reframe(
    index_laspeyres(
      data = .,
      region = "region",
      product = "heading",
      ppp_bh = "sPPP",
      exp_wght = "exp_wght"
    ),
    index_paasche(
      data = .,
      region = "region",
      product = "heading",
      ppp_bh = "sPPP",
      exp_wght = "exp_wght"
    ),
    index_fisher(
      data = .,
      region = "region",
      product = "heading",
      ppp_bh = "sPPP",
      exp_wght = "exp_wght"
    ),
    index_geks(
      data = .,
      region = "region",
      product = "heading",
      ppp_bh = "sPPP",
      exp_wght = "exp_wght"
    )
  ) %>%
  pivot_longer(!c(base_region, region),
    names_to = "index_type",
    values_to = "index_value"
  ) %>%
  mutate(package = "OECDsppps")
```

### Price indices in `pricelevels`

The same index calculations can be carried out with the `pricelevels`
package [@pricelevels], which is built on the `data.table` [@datatable]
library.

```{r}
cpd_oecd_pl <- cpd_oecd %>%
  dplyr::select(heading, region, sPPP, exp_wght) %>%
  as.data.table()

regions_v <- c("area_1", "area_2", "area_3", "area_4", "area_5")

laspeyres_pl <- vector("list", 5)
paasche_pl <- vector("list", 5)
fisher_pl <- vector("list", 5)
geks_pl <- vector("list", 5)

names(laspeyres_pl) <- regions_v
names(paasche_pl) <- regions_v
names(fisher_pl) <- regions_v
names(geks_pl) <- regions_v


for (i in seq_along(regions_v)) {
  laspeyres_pl[[i]] <- cpd_oecd_pl[, laspeyres(
    p = sPPP,
    r = region,
    n = heading,
    w = exp_wght,
    base = c(regions_v[i])
  )]
  paasche_pl[[i]] <- cpd_oecd_pl[, paasche(
    p = sPPP,
    r = region,
    n = heading,
    w = exp_wght,
    base = c(regions_v[i])
  )]
  fisher_pl[[i]] <- cpd_oecd_pl[, fisher(
    p = sPPP,
    r = region,
    n = heading,
    w = exp_wght,
    base = c(regions_v[i])
  )]
  geks_pl[[i]] <- cpd_oecd_pl[, geks(
    p = sPPP,
    r = region,
    n = heading,
    w = exp_wght,
    settings = list(type = "fisher"),
    base = c(regions_v[i])
  )]
}


indices_pricelevels_df <- do.call(cbind, laspeyres_pl) %>%
  as_tibble() %>%
  rownames_to_column(var = "region") %>%
  pivot_longer(!region,
    names_to = "base_region",
    values_to = "laspeyres_index"
  ) %>%
  left_join(
    do.call(cbind, paasche_pl) %>%
      as_tibble() %>%
      rownames_to_column(var = "region") %>%
      pivot_longer(!region,
        names_to = "base_region",
        values_to = "paasche_index"
      ),
    by = c("base_region", "region")
  ) %>%
  left_join(
    do.call(cbind, fisher_pl) %>%
      as_tibble() %>%
      rownames_to_column(var = "region") %>%
      pivot_longer(!region,
        names_to = "base_region",
        values_to = "fisher_index"
      ),
    by = c("base_region", "region")
  ) %>%
  left_join(
    do.call(cbind, geks_pl) %>%
      as_tibble() %>%
      rownames_to_column(var = "region") %>%
      pivot_longer(!region,
        names_to = "base_region",
        values_to = "geks_index"
      ),
    by = c("base_region", "region")
  ) %>%
  pivot_longer(!c(base_region, region),
    names_to = "index_type",
    values_to = "index_value"
  ) %>%
  mutate(
    region = paste("area", region, sep = "_"),
    package = "pricelevels"
  )
```

### Comparing `OECDsppps` and `pricelevels` price indices outputs

The various indices calculated across both packages are numerically
identical.

```{r}
#| fig-height: 12
#| fig-width: 8
indices_oecdsppps_df %>%
  bind_rows(indices_pricelevels_df) %>%
  mutate(
    index_type = str_remove_all(index_type, "_index"),
    base_region = paste("Base region:", base_region)
  ) |>
  ggplot(aes(x = region, y = index_value, fill = package)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Price index comparisons between OECDsppps and pricelevels",
    x = "Region",
    y = "Price index"
  ) +
  theme_minimal() +
  theme(legend.position = "top") +
  scale_fill_manual(values = c("#a3bbdd", "#2a4691")) +
  geom_hline(yintercept = 1) +
  facet_wrap(~ base_region + index_type, ncol = 4)
```

------------------------------------------------------------------------

# Putting it all together: Estimating weighted higher-level aggregates directly from item-level prices {#sec-combined}

As described in the
[implementation](https://amannj.github.io/OECDsppps/articles/Implementation.html)
section, the CPD-GEKS approach is recommended for producing subnational
PPPs by the ICP [@icp2021], and has also been used by a national
statistical institute in a subnational PPP programme, providing
experimental statistics in Italy [@Istat].

This section describes how to implement this [two-step
procedure](https://amannj.github.io/OECDsppps/articles/Implementation.html#fig-aggregation),
which combines:

1.  Estimation of price parities at the basic-heading level using the
    regional extension of the Country-Product-Dummy (CPD) method
    [@summers1973international].
2.  Upon validation, aggregation of BH-level parities into higher-level
    indices using the Gini-Éltetö-Köves-Szulc (GEKS) method, a
    multilateral index construction technique that ensures transitivity,
    in combination with household final consumption expenditure data as
    a weighting structure for household final consumption PPPs
    [@icp2021].

Most importantly, the CPI microdata typically does not contain price
quotes for all COICOP categories, including the ones for which
cross-regional uniform prices can be assumed, such as, for example, for
used cars. However, removing products that are subject to uniform prices
from the estimation will artificially inflate the price variation for
any present category. Consider the example where a generic COICOP class
contains 10 sub-classes, of which 9 are subject to uniform prices while
one sub-class is subject to price variations. If the 9 uniform
sub-classes were to be removed from the sPPPs calculation, the remaining
sub-class with regional price variation would artificially inflate the
price variation of the entire class. It is therefore necessary to
artificially include uniform prices for all sub-aggregates contained
*within* the respective higher aggregate for which sPPPs are estimated.

This can be done with the function `estim_index_link()`, which fills in
missing basic heading PPPs with a value given by the user. The function
returns a data frame containing the variables indicating the region
("region"), basic heading ("product"), basic-heading PPP ("ppp_bh").
Once joined with expenditure weights, this output can be directly fed
into `index_geks()`.[^3]

[^3]: Or any alternative bilateral index formula `index_laspeyres()`,
    `index_paasche()`, `index_fisher()`.

The two steps above can be linked within one pipeline using the
`estim_index_link()` function.

## CPD estimation at the basic-heading level using item-level prices

Basic heading level sPPPs can be obtained from item-level price data by
first grouping the data by basic heading with `group_by()`, then
applying `estim_cpd()` to each basic heading with `group_modify()`.

```{r}
sampledata_prices %>%
  group_by(heading) %>%
  group_modify(~ {
    estim_cpd(.x,
      region = "region",
      product = "product",
      price = "price",
      output = "sPPP"
    )
  }) %>%
  ungroup() %>%
  head() %>%
  gt() %>%
  fmt_number(decimals = 4)
```

This will return a grouped tibble containing three columns: the variable
specified by `group_by()` (in this case the basic heading "group");
region ("region"), and the subnational PPPs estimated via the CPD method
("sPPP").

Note that the returned tibble is still grouped by the grouping variable
specified in `group_by()`, which should be resolved with `ungroup()`
before proceeding further.

## Index calculation using basic-heading indices

GEKS (as well as Laspeyres, Paasche, Fisher, which are all nested within
the GEKS function) indices can be calculated by linking the previous
commands with `index_geks()` (as well as `index_laspeyres()`,
`index_paasche()`, `index_fisher()`) through the `estim_index_link()`
function.

`estim_index_link()` takes a data frame containing the basic heading
sPPPs provided as an input argument `data_sppps`, and joins it with a a
second data frame (supplied via the argument `data_weights`) containing
the household expenditure weights, and gives the user the option to
impute price quotes for expenditure categories, for which household
expenditure weights are present but nor price quotes are available.

Thus, `estim_index_link()` returns a data object with all necessary
variables for the calculation of the price indices, e.g.,
`index_geks()`.

To summarise, the complete workflow has these three steps:

1.  Grouping item-level price data by basic heading, estimating basic
    heading sPPPs, and ungrouping the output data
2.  Preparing the data with `estim_index_link()` for index number
    calculation
3.  Estimating the final index numbers with `index_laspeyres()`,
    `index_paasche()`, `index_fisher()`, and `index_geks()`

```{r}
sampledata_prices %>%
  # 1st step: grouping the data
  group_by(heading) %>%
  # 1st step: estimating the basic-heading sPPPs
  group_modify(~ {
    estim_cpd(.x,
      region = "region",
      product = "product",
      price = "price",
      output = "sPPP"
    )
  }) %>%
  # 1st step: ungrouping the output data
  ungroup() %>%
  # 2nd step: preparing the data for index number calculation
  estim_index_link(
    data_sppps = .,
    data_weights = sampledata_weights,
    product_heading = "heading",
    region = "region",
    sPPP = "sPPP",
    weights = "exp_wght",
    complete_sppp = NA
  ) %>%
  # 3rd step: Index number calculation
  index_geks() %>%
  # Previewing the results
  head() %>%
  gt() %>%
  fmt_number(decimals = 4)
```

## Incomplete input data

Unfortunately, the required input data is often incomplete and requires
further processing before it can be used. Item-level price data can be
validated through `valid_pot()`, `valid_apt()`, `valid_ratio_xr()`, and
`valid_ratio_ppp()`. Expenditure weights can be validated through
`valid_est()`.

### Missing price quotes

Basic heading sPPPs obtained from `estim_cpd()` may be incomplete for
several reasons, such as missing price quotes for specific regions or
uniform pricing across regions. In these cases, the above-outlined
approach results in an error from `index_geks()` due to an incomplete
sPPP matrix.

In the example below, we remove the sPPPs for two region/basic-heading
pairs. While `estim_index_link()` returns a harmonised data frame,
`index_geks()` returns an error.

Note that in the code chunk below, `tryCatch()` is used to print the
error message returned by `index_geks()` as an output.

```{r}
# Return error message ------
return_error <- function(expr) {
  tryCatch(expr,
    error = function(e) {
      message("Error message:\n", e)
    }
  )
}


return_error(
  sampledata_prices %>%
    group_by(heading) %>%
    group_modify(~ {
      estim_cpd(.x,
        region = "region",
        product = "product",
        price = "price",
        output = "sPPP"
      )
    }) %>%
    ungroup() %>%
    # Remove sPPPs for the following region/heading combinations: 1/1, 2/1
    filter(!(region %in% c("region_1", "region_2") & heading == "heading_1")) %>%
    estim_index_link(
      data_sppps = .,
      data_weights = sampledata_weights,
      product_heading = "heading",
      region = "region",
      sPPP = "sPPP",
      weights = "exp_wght",
      complete_sppp = NA
    ) %>%
    index_geks()
)
```

`estim_index_link()` enables imputing a user-specified value for missing
sPPPs through the "complete_sppp" argument. The same value is imputed
for all region/basic-heading pairs with missing sPPPs. The missing pairs
are identified by `estim_index_link()` and do not have to be specified
by the user. Below, we impute 1 for sPPPs of all such pairs.
`estim_index_link()` also returns a warning listing all region/basic
heading pairs for which the user-provided value was imputed.

```{r}
sampledata_prices %>%
  group_by(heading) %>%
  group_modify(~ {
    estim_cpd(.x,
      region = "region",
      product = "product",
      price = "price",
      output = "sPPP"
    )
  }) %>%
  ungroup() %>%
  # Remove sPPPs for the following region/heading combinations: 1/1, 2/1
  filter(!(region %in% c("region_1", "region_2") & heading == "heading_1")) %>%
  # Setting complete_sppp = 1 to impute 1 for missing sPPPs
  estim_index_link(
    data_sppps = .,
    data_weights = sampledata_weights,
    product_heading = "heading",
    region = "region",
    sPPP = "sPPP",
    weights = "exp_wght",
    complete_sppp = 1
  ) %>%
  index_geks() %>%
  head() %>%
  gt() %>%
  fmt_number(decimals = 4)
```

### Incomplete weighting matrix

Alternatively, the expenditure weights data frame can contain empty
fields for some region/basic-heading combinations. `estim_index_link()`
still works with such data and returns a standardised data frame.
However, this data frame contains NAs for some region/basic-heading
weights.

```{r}
# Remove expenditure weight for the following region/heading combination: 1/1
sampledata_weights <- sampledata_weights %>%
  filter(!(heading %in% c("heading_1") & region %in% c("region_1")))

sampledata_prices %>%
  group_by(heading) %>%
  group_modify(~ {
    estim_cpd(.x,
      region = "region",
      product = "product",
      price = "price",
      output = "sPPP"
    )
  }) %>%
  ungroup() %>%
  estim_index_link(
    data_sppps = .,
    data_weights = sampledata_weights,
    product_heading = "heading",
    region = "region",
    sPPP = "sPPP",
    weights = "exp_wght",
    complete_sppp = NA
  ) %>%
  head() %>%
  gt() %>%
  fmt_number(decimals = 4)
```

In turn, this causes the index number function to break and return an
error due to an incomplete expenditure weights matrix.

Note that in the code chunk below, `tryCatch()` is used to print the
error message returned by `index_geks()` as an output.

```{r}
return_error(
  sampledata_prices %>%
    group_by(heading) %>%
    group_modify(~ {
      estim_cpd(.x,
        region = "region",
        product = "product",
        price = "price",
        output = "sPPP"
      )
    }) %>%
    ungroup() %>%
    estim_index_link(
      data_sppps = .,
      data_weights = sampledata_weights,
      product_heading = "heading",
      region = "region",
      sPPP = "sPPP",
      weights = "exp_wght",
      complete_sppp = NA
    ) %>%
    index_geks()
)
```

<br> <br>

------------------------------------------------------------------------

# References {.unnumbered}
