---
title: "Validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vignette}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    number-depth: 5    
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(gt)
library(ggplot2)
library(OECDsppps)
```

# Item-level validation

The documentation for data validation is illustrated using a snippet of the
[UK CPI microdata](https://www.ons.gov.uk/economy/inflationandpriceindices/datasets/consumerpriceindicescpiandretailpricesindexrpiitemindicesandpricequotes).

```{r}
head(uk_cpi, n = 3) |> gt()
```

## Intra- and inter-regional/country validation

-   Two step procedure following international recommendations [@icp2021, page 26] and [@worldbank2013, p.249], trying to identify and eliminate non-sampling,[^1] and production errors in the underlying data.
-   Intra-regional validation: prices collected are edited and verified *within* a region or territorial area
-   Inter-region validation: prices validation *across all* regions/countries, ensuring that average prices are based on comparable products across countries and that products have been accurately priced.

[^1]: Non-sampling errors occur for reasons such as pricing the wrong product (product error) or incorrectly recording the product’s price or unit of measure (pricing error).

### Intra-regional validation

-   Intra-region validation establishes that price collectors within the same region have priced products that match the product specifications and that the prices they have reported are correct. This is done in two stages, which correspond to the outlier detection of (a) individual prices and (b) average price aggregates.

#### Stage 1: individual price outlier statistcs

-   For each product, a **Price Observation Table** is obtained, containing a characterisation of the individual product as well as two *individual price outlier statistics*, the *Ratio-to-average price test* and the *t-value test*; see @worldbank2013, table 9.1a for an extensive example.

-   *Ratio-to-average price test*: The ratio of an individual price observation $i$, $P_{i}$, of a specific product $j$ and the observed average price for the product, $\mu_j$. An observed price passes the this test if the ratio is between 0.5 and 1.5, meaning it is no less than half or no more than double the mean price. This simple check flags extreme values without relying on standard deviation, which can itself be distorted by outliers [@worldbank2013, p.251].

$$
ratio-to-average = p_{ij}/\mu_j
$$

-   *t-value test*: The ratio of the deviation of an individual price observation from the average reference quantity price for the product and the standard deviation of the product. To pass the test, the ratio must be 2.0 or less (any value greater than 2.0 is suspect because it falls outside the 95 percent confidence interval).

$$
t-val = (p_{ij} - \mu_{P_j}) / \sigma_{P_j}
$$

-   Individual price quotes that do not pass these tests are flagged in the price observation table. The two statistics are implemented in the OECD price statistics package `OECD_SPPP`. The price observation table is generated with the function `valid_pot()`, which calls `valid_r2a` and `valid_tval`.

------------------------------------------------------------------------

**Example using UK CPI microdata**


```{r}
#Price Observation Table
uk_pot <- uk_cpi |>
  valid_pot()

head(uk_pot, n = 3) |>
  gt() |>
  tab_header(
    title = md("**Price Observation Table**"),
    subtitle = md("Example for `item_id` = 210111")
  ) |>
  fmt_number(
    columns = c(
      `Ratio-to-average price test`,
      `T-value test`
    ),
    decimals = 1
  )

uk_pot |>
  select(
    `Product code`,
    `Ratio-to-average price test`:`T-value test`
  ) |>
  pivot_longer(`Ratio-to-average price test`:`T-value test`) |>
  mutate(
    is.outlier = case_when(name == "Ratio-to-average price test" & (value < 0.5 | value > 1.5) ~ "Test not passed",
      name == "T-value test" & (value > 2) ~ "Test not passed",
      .default = "Test passed"
    ),
    is.outlier = factor(is.outlier, levels = c("Test passed", "Test not passed"))
  ) |>
  ggplot(aes(x = value, fill = is.outlier)) +
  facet_wrap(~name, scales = "free") +
  geom_histogram() +
  labs(
    title = "Individual price outlier statistcs",
    subtitle = "White sliced loaf branded 750G, 2018",
    x = "",
    fill = ""
  ) +
  theme_minimal()+
  theme(legend.position = "top")
```

#### Stage 2: Average price statistics

-   This stage involves identifying extreme values among the average prices of the products listed in the average price table. An extreme value is defined as an individual price or average price that for a given test scores a value that falls outside a predetermined critical value and is build on two *average price outlier statistics*, which are summarised in the **Average Price table**; see @worldbank2013, table 9.2a and 9.2b for an extensive example. The two statisticis contained in this table are the *max-min ratio test* and the *coefficient to variation* test.

-   *max-min ratio test*: The ratio between the maximal and minimal observed price for product $j$. Products where the maximal observed price is more than twice as big as the minimum are flagged

$$
max-min~ratio = max(p_j)/min(p_j)
$$

-   *coefficient to variation test*: The standard deviation for the product expressed as a percentage of the average price for the product. Products with a coefficient of variation greater than 20% will be flagged.

$$
coefficient-to-variation: \sigma_{p_j} / \mu_{p_j}
$$

-   Product price quotes that do not pass these tests are flagged in the *average price table* and is created using function `valid_apt` in the OECD_SPPPs package.

------------------------------------------------------------------------

**Example using UK CPI microdata**

```{r}
# Average Price Table
uk_apt <- uk_cpi |>
  valid_apt()

head(uk_apt) |>
  gt() |>
  tab_header(
    title = md("**Average Price Table**"),
    subtitle = md("Example for `item_id` = 210111, **pre-cleaning**")
  ) |>
  fmt_number(columns = -c(Year, `Product code`),decimals = 1)
```

#### Linking validation pipelines

-   The degree of validation depends on the quality of the underlying microdata. With unconsolidated and/or raw data, more extensive revisions of the raw data may be necessary.
    -   For example, these would require the validation and checking of the reference quantity price.
    -   For the selected product, a bimodal distribution is observed (after removing outliers).
-   Eliminating these makes selected item pass all intra-regional validation checks.

```{r}
uk_irv <- uk_cpi |>
  # Apply Price Observation Table checks
  valid_pot() |>
  # Condition on price quotes which pass the Price Observation Table tests
  filter(!`Ratio-to-average price test FLAG` & !`T-value test FLAG`) |>
  # Remove bimodal distribution
  filter(`Ratio-to-average price test` > 0.8) |>
  select(-c(contains("Ratio-to-average"), contains("T-value"))) |>
  # Apply Average Price Table checks
  valid_apt()

uk_irv |>
  gt() |>
  tab_header(
    title = md("**Average Price Table**"),
    subtitle = md("Example for `item_id` = 210111, **post-cleaning**")
  ) |>
  fmt_number(
    columns = -c(Year, `Product code`),
    decimals = 1
  )
```

### Inter-regional validation

...


# References
